---
title: "Exercise on troubleshooting: Solution"
author: "Radchuk Viktoriia"
date: "2022-1-22"
output:
  html_document:
    theme: default
       
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(lattice)
library(glmmTMB)

source(file = here::here('data', 'Prep_data.R'))   ## consider leaving out and directly loading df.Rda

## load data on metabolic rates - in case above line does not work
# load(file = here::here('data', 'df_metab_Dormice.Rda'))
```

# One possible solution for the exercise

We work with the data on hibernating dormice

We rescale the data for better model convergence. Feel free to try and fit the model without this step (I already did that) and see what happens. 
``` {r model arousal prob, echo = TRUE, fig.align = 'center'}
df <- df[! (is.na(df$HoursBegin) | is.na(df$Tamb)), ]
min(df$HoursBegin)

# coding year as a factor
df$Y <- 2009
df$Y[df$Year %in% c(2011, 2012)] <- 2011
df$Y <- as.factor(df$Y)

### rescaling time since hibernation and temperature for better convergence
df$HoursScale <- (df$HoursBegin - mean(df$HoursBegin, na.rm = T)) / sd(df$HoursBegin, na.rm = T)
df$HoursScale2 <- df$HoursScale^2
hist(df$HoursScale, 
     xlab = 'Standarized time since begin hibernation',
     main = NULL, col = 'black')

df$T_scale <- (df$Tamb - mean(df$Tamb, na.rm = T)) / sd(df$Tamb, na.rm = T)
df$Tscale2 <- df$T_scale ^ 2
hist(df$T_scale, xlab = 'Standardized temperature', 
     main = NULL, col ='black')
```


## Fit a mixed-effects Binomial model 
This is the model with autocorrelation, a quadratic effect of temperature, a linear effect of time since the beginning of hibernation, and an effect of year
```{r noQuad, echo = TRUE}
df$Time_fac <- as.factor(df$Timecon)

modRandAR_no2Hour_prob <- glmmTMB(state ~ HoursScale + 
                          T_scale + Tscale2 +
                          Y + (1 | ID) +  ar1(Time_fac - 1|ID), 
                         family = binomial,
                         data=df)
summary(modRandAR_no2Hour_prob)
```

## One way to solve the convergence issue 
By using a different optimiser, we go with 'BFGS'
```{r convergence issue, echo = TRUE}
mod_BFGS <- update(modRandAR_no2Hour_prob, control=glmmTMBControl(optimizer=optim,
                                       optArgs=list(method="BFGS")))
summary(mod_BFGS)
```


# Questions?


