---
title: "Solution for the exercise on mixed-effect models"
author: "Radchuk Viktoriia"
date: "2022-1-2"
output:
  html_document:
    theme: default
       
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(lme4)
library(spaMM)
library(merTools)
```

# Read in the data

```{r read-data, echo = TRUE}
dat <- read.csv(file = here::here('data', 'dat_phen_temp_subs.csv'))
str(dat)
```


# Data preparation
```{r data prep, echo = TRUE, fig.align='center'}
dat_fac <- dat %>%
  mutate_if(is.character, as.factor) %>% 
  mutate(ID = as.factor(ID))
str(dat_fac)

```

# Relation between temperature and phenology per species

```{r relation per species, echo = TRUE, out.width= '100%', fig.align='center'}
ggplot(dat_fac, aes(x = det_Clim, y = Trait_mean)) +
  geom_point() + theme_bw() + 
  facet_wrap(vars(ID)) +
  xlab('Standardized temperature') +
  ylab('Standardized phenology')

```

# Fit mixed effects model with random intercept using `lme4`
```{r mod-randInt, echo= TRUE}
mod_randInt <- lmer(Trait_mean ~ det_Clim + (1|Species), data = dat_fac, REML = FALSE)
summary(mod_randInt)

mod_randInt_spam <- fitme(Trait_mean ~ det_Clim + (1|Species), data = dat_fac, method = 'ML')
summary(mod_randInt_spam)
```
Pay attention: singularity! What does it mean?


# Predictions with the fitted model
```{r Pred randInterc, echo = TRUE, out.width = '100%', fig.align='center'}

new <- expand.grid('Species' = unique(dat_fac$Species), 
                       det_Clim = seq(from = min(dat_fac$det_Clim), to = 
                                    max(dat_fac$det_Clim), length.out = 25))

head(new)

new$pred_fix <- predict(mod_randInt, newdata = new, re.form = NA)
new$pred <- predict(mod_randInt, newdata = new) ## by default re.form = NULL, i.e. including all random effects
ggplot(data = new, aes(x = det_Clim, y = pred, group = Species, col = Species)) +
  geom_line(lty = 3) + theme_bw() +
  geom_line(data = new, aes(x = det_Clim, y = pred_fix), col = 'black') +
  xlab('Standardized temperature') + ylab('Standardized phenology')
```


# Add 95%CI ot the plotted fit
We use the `predictInterval()` function from the `merTools` package to predict the CI around each predicted value.   
```{r Pred randInterc CI, echo = TRUE, out.width = '100%', fig.align='center'}

CI_pred <- predictInterval(mod_randInt, new, level = 0.95)
CI_pred_fix <- predictInterval(mod_randInt, new, level = 0.95, which = 'fixed', include.resid.var = FALSE)
CI_pred_fix <- CI_pred_fix %>% 
  rename(upr_Fix = upr, lwr_Fix = lwr) %>% 
  dplyr::select(., !fit)

## important: we here first plot the intervals (using geom_ribbon()) for all species, and only then the line (otherwise it iwll be overplotted and not visible)
new_full <- cbind(new, CI_pred, CI_pred_fix)
ggplot(data = new_full, aes(x = det_Clim, y = pred_fix, group = Species)) +
  geom_ribbon(data = new_full, aes(x = det_Clim, ymin = lwr, ymax = upr), fill = 'grey', alpha = 0.3) +
  geom_ribbon(data = new_full, aes(x = det_Clim, ymin = lwr_Fix, ymax = upr_Fix), fill = 'blue',  alpha = 0.3) +
  geom_line(lty = 3) + theme_bw() +
  xlab('Standardized temperature') + ylab('Standardized phenology')
```


# Fit mixed-effects model with random slope using `lme4`
```{r mod-randSlope, echo= TRUE}
mod_randSlope <- lmer(Trait_mean ~ det_Clim + (1 + det_Clim|Species), data = dat_fac, REML = FALSE)
summary(mod_randSlope)

mod_randSl_spam <- fitme(Trait_mean ~ det_Clim + (1 + det_Clim|Species), data = dat_fac, method = 'ML')
summary(mod_randSl_spam)
```
Again: singularity when using `lme4`!   Would we use these models to make any predictions?? 


# Predictions with the fitted random-slope model
```{r Pred randSl, echo = TRUE, out.width = '100%', fig.align='center'}

newSlope <- expand.grid('Species' = unique(dat_fac$Species), 
                       det_Clim = seq(from = min(dat_fac$det_Clim), to = 
                                    max(dat_fac$det_Clim), length.out = 25))

head(newSlope)

newSlope$pred_fix <- predict(mod_randSlope, newdata = newSlope, re.form = NA)
newSlope$pred <- predict(mod_randSlope, newdata = newSlope) ## by default re.form = NULL, i.e. including all random effects
ggplot(data = newSlope, aes(x = det_Clim, y = pred, group = Species, col = Species)) +
  geom_line(lty = 3) + theme_bw() +
  geom_line(data = newSlope, aes(x = det_Clim, y = pred_fix), col = 'black') +
  xlab('Standardized temperature') + ylab('Standardized phenology')
```


# Add 95%CI ot the plotted fit
We use the `predictInterval()` function from the `merTools` package to predict the CI around each predicted value.   
```{r Pred randSlope CI, echo = TRUE, out.width = '100%', fig.align='center'}

CI_pred <- predictInterval(mod_randSlope, newSlope, level = 0.95)  ## note warnings!!!!
CI_pred_fix <- predictInterval(mod_randSlope, newSlope, level = 0.95, which = 'fixed', include.resid.var = FALSE)  ## note warnings!
CI_pred_fix <- CI_pred_fix %>% 
  rename(upr_Fix = upr, lwr_Fix = lwr) %>% 
  dplyr::select(., !fit)

## important: we here first plot the intervals (using geom_ribbon()) for all species, and only then the line (otherwise it will be overplotted and not visible)
newSlope <- cbind(newSlope, CI_pred, CI_pred_fix)
ggplot(data = newSlope, aes(x = det_Clim, y = pred_fix, group = Species)) +
  geom_ribbon(data = newSlope, aes(x = det_Clim, ymin = lwr, ymax = upr), fill = 'grey', alpha = 0.3) +
  geom_ribbon(data = newSlope, aes(x = det_Clim, ymin = lwr_Fix, ymax = upr_Fix), fill = 'blue',  alpha = 0.3) +
  geom_line(lty = 3) + theme_bw() +
  xlab('Standardized temperature') + ylab('Standardized phenology')
```




