---
title: "Solution: testing hypothesis 2 with the data on dormice"
author: "Radchuk Viktoriia"
date: "2022-1-5"
output:
  html_document:
    theme: default
       
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(glmmTMB)
library(car)
library(DHARMa)

source(file = here::here('data', 'Prep_data.R'))   
```

# Hypothesis
- No effect of temperature on metabolic rate of normothermic individuals is expected because they should be able to thermoregulate


# Read in and prepare the data


```{r data prep, echo = TRUE}
df$Log_MR <- log(df$MR_h)
df$Tamb2 <- df$Tamb^2
df$Y <- 2009
df$Y[df$Year %in% c(2011, 2012)] <- 2011
df$Y <- as.factor(df$Y)
df$ID <- as.factor(df$ID)
```


``` {r split per state, echo = TRUE, fig.align = 'center'}
## just use the subset of the data with the individuals in state == 1 (aroused)
aroused <- df[df$state == 1, ]

nrow(aroused)  ### 2765

hist(aroused$Month)  
xyplot(Log_MR ~ Tamb|ID, data = aroused, xlab = expression('Ambient temperature, ' ~degree~C), ylab = 'Log metabolic rate')

# exclude missing values 
aroused <- aroused[!is.na(aroused$Log_MR), ]
```

## Mixed-effects model with ID as random intercept
I will use glmmTMB here but you may have chosen a different R package 
``` {r  arousal ME randI, echo = TRUE}
modRandID_arous <- glmmTMB(Log_MR ~ Tamb  + Y + (1|ID), 
                        data = aroused)
summary(modRandID_arous)
```

## Model diagnostics 
Using DHARMa directly.  

```{r moddiagn RInt, echo = TRUE, fig.align='center'}
sim <- simulateResiduals(modRandID_arous, plot = T)  
testUniformity(sim, plot = F)
testQuantiles(sim, plot = T)

# plot residuals vs our predictors
plotResiduals(sim, form = aroused$Tamb)
plotResiduals(sim, form = aroused$Y)
```
<span style="color:orange"> How do we interpret these plots? </span>

Let us test for temporal autocorrelation, since the measurements were taken over time for each individual.     
```{r autocor RIntMod, echo = TRUE, fig.align='center'}
sim_ID10 <- recalculateResiduals(sim, sel = aroused$ID == 10)
testTemporalAutocorrelation(sim_ID10, time = aroused$Timecon[aroused$ID == '10'])

```


```{r fun testAutocor, echo = FALSE}
fun_autocor <- function(simulatedObj, data, ID){
sim_subset <- recalculateResiduals(simulatedObj, sel = data$ID == ID)
testAut <- testTemporalAutocorrelation(sim_subset, time = data$Timecon[data$ID == ID], plot = F)
return(list(testAut, data.frame(DurbW = testAut$statistic, p = testAut$p.value, ID = ID)))  
}

```

Using the same function we used in the lecture to test temporal autocorrelation for each individual in the dataset. 
```{r use fun testAutocor, echo = TRUE, fig.align='center', out.width= '60%'}
levels(aroused$ID)
fun_autocor(sim, aroused, ID = '10')

```


### Test autocorrelation in all studies
```{r testAutocor All Stud, echo = TRUE}
dat_autoc <- data.frame(DurbW =numeric(), p = numeric())
dat_autoc <- rbind(dat_autoc, fun_autocor(sim, aroused, 10)[[2]])
for(i in unique(aroused$ID)[-1]){
  dat_autoc <- rbind(dat_autoc, fun_autocor(sim, aroused, ID = i)[[2]])
}

length(levels(aroused$ID))  
dat_autoc$sign <- ifelse(dat_autoc$p < 0.05,'Yes','No')
sum(dat_autoc$sign == 'Yes')
sum(dat_autoc$sign == 'Yes')/length(levels(aroused$ID))
```
So, temporal autocorrelation is an issue here.    


## Mixed-effects model accounting for temporal autocorrelation
``` {r  arousal ME randI autocor, echo = TRUE}
aroused$Time_fac <- as.factor(aroused$Timecon)
modRandID_arousAR <- glmmTMB(Log_MR ~ Tamb + Y + (1|ID) +
                             ar1(Time_fac - 1|ID), 
                        data = aroused)
summary(modRandID_arousAR)
```

## Model diagnostics 
Remember we need to rotate the residuals otherwise there is no point looking at the diagnostics (the autocorrealtion will still be there in the residual structure)

This takes long time, run it in your free time.
```{r moddiagn RIntAR1, echo = TRUE, eval = FALSE}
## Do not run
simAR1_rot <- simulateResiduals(modRandID_arousAR, plot = T, rotation = 'estimated')
testUniformity(simAR1_rot, plot = F)
testQuantiles(simAR1_rot, plot = T)
# plot residuals vs our predictors
plotResiduals(simAR1_rot, form = aroused$Tamb)
plotResiduals(simAR1_rot, form = aroused$Y)
```
<span style="color:orange"> How do we interpret these plots? </span>

# Model inference
Interpreting the results of the model  (assuming that the model diagnostics were fine, first make sure to run the diagnostics in full!).    
We use LRT to test the significance of the linear effect of temperature on metabolic rate in normothermic individuals. 

``` {r  torpor LRT ML, echo = TRUE}
modRandID_arous_noT <- glmmTMB(Log_MR ~ Y + (1|ID) +
                                 ar1(Time_fac - 1|ID), 
                        data = aroused)
summary(modRandID_arous_noT)

## pay attention: we by default fitted models with ML so we can run LRT to test the significance of the fixed effect

anova(modRandID_arous_noT, modRandID_arousAR)  ## significant effect
```
<span style="color:orange"> Do we find support for our hypothesis? </span>

# Plot predictions
```{r newDat, echo = TRUE, fig.align='center'}
min(aroused$Tamb)
new_dat <- expand.grid('ID' = unique(aroused$ID), 
                       'Y' = unique(aroused$Y),
                       'Tamb' = seq(from = min(aroused$Tamb), to = 
                                    max(aroused$Tamb), length.out = 20))

head(new_dat)

new_dat$Time_fac <- as.factor(rep('10', nrow(new_dat)))  ## keeping just to one time point
new_dat$pred_fixTMB <- predict(modRandID_arousAR, newdata = new_dat, re.form = NA, type = 'response')

ggplot(data = new_dat, aes(x = Tamb, y = pred_fixTMB)) +
  geom_point() + theme_bw() +
  facet_grid(cols = vars(Y)) +
  xlab('Ambient temperature') + ylab('Metabolic rate')
```

Adding confidence intervals to the plots
```{r pred with CI, echo = TRUE, fig.align='center'}
new_dat$pre_SE_fixTMB <- predict(modRandID_arousAR, newdata = new_dat, re.form = NA, se.fit = TRUE, type = 'response')$se.fit 
new_dat$pred_upper_fix <- new_dat$pred_fixTMB + qnorm(0.975)*new_dat$pre_SE_fixTMB 
new_dat$pred_lower_fix <- new_dat$pred_fixTMB - qnorm(0.975)*new_dat$pre_SE_fixTMB 
head(new_dat)
ggplot(data = new_dat, aes(x = Tamb, y = pred_fixTMB)) +
  theme_bw() + facet_grid(cols = vars(Y)) +
  geom_ribbon(data = new_dat, aes(x = Tamb, ymin = pred_lower_fix, ymax = pred_upper_fix), alpha = 0.4) +
  geom_line(lty  = 3) +
    xlab('Ambient temperature') + ylab('Metabolic rate')
```


